Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

### 2025-08-29 manuscript.tex
fixing some typos

\abstract{Time-resolved electrophysiological measurements such as those
obtained through magneto- and electroencephalography (M/EEG) offer a
unique window onto the neural activity underlying cognitive processes.
Researchers are often interested in determining whether and when these
signals differ across experimental conditions or participant groups. The
conventional approach involves mass-univariate statistical testing
across time and space followed by corrections for multiple comparisons
or some form of cluster-based inference. While effective for controlling
error rates at the cluster-level, cluster-based inference comes with a
significant limitation: by shifting the focus of inference from
individual time points to clusters, it prevents drawing conclusions
about the precise onset or offset of observed effects. Here, we present
a \emph{model-based} alternative for analysing M/EEG timeseries, such as
event-related potentials or time-resolved decoding accuracy. Our
approach leverages Bayesian generalised additive multilevel models,
providing posterior odds that an effect exceeds zero (or chance) at each
time point, while naturally accounting for temporal dependencies and
between-subject variability. Using both simulated and empirical M/EEG
datasets, we show that this approach substantially outperforms
conventional methods in estimating the onset and offset of neural
effects, yielding more precise and reliable estimates. We provide an
open-source R package implementing the method and describe how it can be
integrated into M/EEG analysis pipelines using MNE-Python. }

@@

Understanding the temporal dynamics of cognitive processes requires
methods that can capture fast-changing neural activity with high
temporal resolution. Magnetoencephalography and electroencephalography
(M/EEG) are two such methods, widely used in cognitive neuroscience for
their ability to track brain activity at the millisecond scale. These
techniques provide rich time series data that reflect how neural
responses unfold in response to stimuli or tasks. A central goal in many
M/EEG studies is to determine whether, when, and where neural responses
differ across experimental conditions or groups.

The conventional approach involves mass-univariate statistical testing
through time and/or space followed by some form of correction for
multiple comparisons with the goal of maintaining the familywise error
rate (FWER) or the false discovery rate (FDR) at the nominal level
(e.g., 5\%). Cluster-based inference is the most common way of achieving
this sort of error control in the M/EEG literature, being the
recommended approach in several software programs (e.g.,
\texttt{EEGlab}, \citeproc{ref-delorme2004}{Delorme \& Makeig, 2004};
\texttt{MNE-Python}, \citeproc{ref-gramfort2013}{Gramfort, 2013}). While
effective for controlling error rates, cluster-based inference comes
with a significant limitation: by shifting the focus of inference from
individual datapoints (e.g., timesteps, sensors, voxels) to clusters, it
prevents the ability to draw precise conclusions about the
spatiotemporal localisation of such effects
(\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}). As
pointed out by Maris \& Oostenveld (\citeproc{ref-maris2007}{2007}):
``there is a conflict between this interest in localized effects and our
choice for a global null hypothesis: by controlling the FA {[}false
alarm{]} rate under this global null hypothesis, one cannot quantify the
uncertainty in the spatiotemporal localization of the effect''. Even
worse, Rosenblatt et al. (\citeproc{ref-rosenblatt2018}{2018}) note that
cluster-based inference suffers from low spatial resolution: ``Since
discovering a cluster means that `there exists at least one voxel with
an evoked response in the cluster', and not that `all the voxels in the
cluster have an evoked response', it follows that the larger the
detected cluster, the less information we have on the location of the
activation.'' As a consequence, cluster-based inference is expected to
perform poorly for identifying the onset of M/EEG effects; a property
that was later demonstrated in simulation studies (e.g.,
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}).

@@

Scalp-recorded M/EEG signals capture neural activity originating from
various brain regions and are often contaminated by artifacts unrelated
to the cognitive processes under investigation. Consequently, analysing
M/EEG data necessitates methods that can disentangle task-relevant
neural signals from extraneous ``noise.'' A widely adopted technique for
this purpose is the estimation of event-related potentials (ERPs), which
are stereotyped electrophysiological responses time-locked to specific
sensory, cognitive, or motor events. Typically, ERPs are derived by
averaging EEG or MEG epochs across multiple trials aligned to the event
of interest (e.g., stimulus onset), thereby enhancing the
signal-to-noise ratio by attenuating non-time-locked activity. However,
this averaging approach has notable limitations: it assumes consistent
latency and amplitude across trials and is primarily suited for simple
categorical designs. Such assumptions may not hold in more complex
experimental paradigms, potentially leading to suboptimal ERP
estimations (e.g., \citeproc{ref-smith2014a}{N. J. Smith \& Kutas,
2014a}).

To overcome the limitations of simple averaging, several model-based
approaches for estimating ERPs have been proposed. These methods are
motivated by the observation that traditional ERP averaging is
mathematically equivalent to fitting an intercept-only linear regression
model in a simple categorical design without overlapping events
(\citeproc{ref-smith2014a}{N. J. Smith \& Kutas, 2014a}). In contrast to
simple averaging, regression-based approaches to ERP estimation offer
substantially greater flexibility. Notably, they allow for the modelling
of both linear and nonlinear effects of continuous predictors, such as
word frequency or age (e.g., \citeproc{ref-smith2014a}{N. J. Smith \&
Kutas, 2014a}, \citeproc{ref-smith2014b}{2014b};
\citeproc{ref-tremblay2014}{Tremblay \& Newman, 2014}), and enable the
disentangling of overlapping cognitive processes (e.g.,
\citeproc{ref-ehinger_unfold_2019}{Ehinger \& Dimigen, 2019};
\citeproc{ref-skukies_brain_2024}{Skukies et al., 2024};
\citeproc{ref-skukies_modelling_2021}{Skukies \& Ehinger, 2021}). One
widely used implementation of this approach is provided by the LIMO EEG
toolbox (\citeproc{ref-pernet2011}{Pernet et al., 2011}), which follows
a multi-stage analysis pipeline. First, a separate regression model is
fit for each datapoint (e.g., each time point and electrode) at the
individual participant level to estimate ERP responses. This is followed
by group-level statistical analyses of the resulting regression
coefficients, often accompanied by corrections for multiple comparisons
or cluster-based inference (for recent applied examples, see
\citeproc{ref-dunagan2025}{Dunagan et al., 2025};
\citeproc{ref-wuxfcllhorst2025}{Wüllhorst et al., 2025}).

@@

Beyond modelling nonlinear effects of continuous predictors on ERP
amplitudes, GAMs have been employed to capture the temporal dynamics of
ERPs themselves, effectively modelling the shape of the waveform over
time (\citeproc{ref-abugaber2023}{Abugaber et al., 2023};
\citeproc{ref-baayen2018}{Baayen et al., 2018};
\citeproc{ref-meulman2015}{Meulman et al., 2015},
\citeproc{ref-meulman2023}{2023}). This approach allows for the
estimation of smooth, data-driven functions that characterise how neural
responses evolve over time, offering a flexible alternative to
traditional linear models. In the following section, we provide a brief
introduction to GAMs, highlighting their applicability to M/EEG time
series analysis and the advantages they offer over conventional methods.

@@

A detailed treatment of the technical underpinnings of GAMs is beyond
the scope of this article (see reference books such as
\citeproc{ref-hastie2017}{Hastie \& Tibshirani, 2017};
\citeproc{ref-wood2017}{Wood, 2017a}). However, it is worth emphasising
that GAMs have been successfully applied to a wide range of time series
data across the cognitive sciences, including pupillometry (e.g.,
\citeproc{ref-vanrij2019}{Rij et al., 2019}), articulography (e.g.,
\citeproc{ref-wieling2018}{Wieling, 2018}), speech formant dynamics
(e.g., \citeproc{ref-suxf3skuthy2021}{Sóskuthy, 2021}), neuroimaging
data (e.g., \citeproc{ref-dinga2021}{Dinga et al., 2021}), and
event-related potentials (e.g., \citeproc{ref-abugaber2023}{Abugaber et
al., 2023}; \citeproc{ref-baayen2018}{Baayen et al., 2018};
\citeproc{ref-meulman2015}{Meulman et al., 2015},
\citeproc{ref-meulman2023}{2023}). Their appeal for modelling M/EEG data
lies in their ability to flexibly capture the complex shape of ERP
waveforms without overfitting, through the use of smooth functions
constrained by penalisation. Recent extensions, such as distributional
GAMs (\citeproc{ref-rigby2005}{Rigby \& Stasinopoulos, 2005};
\citeproc{ref-umlauf2018}{Umlauf et al., 2018}), allow researchers to
model not only the mean structure but also the variance (or scale) and
other distributional properties as functions of predictors, a feature
that has proven useful in modelling neuroimaging data (e.g.,
\citeproc{ref-dinga2021}{Dinga et al., 2021}). Moreover, hierarchical or
multilevel GAMs (\citeproc{ref-pedersen_hierarchical_2019}{E. J.
Pedersen et al., 2019}) provide a principled way to account for the
nested structure of M/EEG data (e.g., trials within participants),
enabling the inclusion of varying intercepts, slopes, and smoothers (as
illustrated in Figure~\ref{fig-intro-gam}\ignorespaces C-D). This
approach mitigates the risk of overfitting and reduces the influence of
outliers on smooth estimates (\citeproc{ref-baayen2020}{Baayen \& Linke,
2020}; \citeproc{ref-meulman2023}{Meulman et al., 2023}).

@@

We then compared the ability of the BGAM to accurately estimate the
onset and offset of the ERP difference to other widely-used methods.
First, we conducted mass-univariate t-tests (thus treating each timestep
independently) and identified the onset and offset of the ERP difference
as the first and last values crossing an arbitrary significance
threshold (\(\alpha = 0.05\)). We then followed the same approach but
after applying different forms of multiplicity correction to the
\(p\)-values. We compared two methods that control the FDR (i.e.,
\texttt{BH95}, \citeproc{ref-benjamini1995}{Benjamini \& Hochberg,
1995}; and \texttt{BY01}, \citeproc{ref-benjamini2001}{Benjamini \&
Yekutieli, 2001}), one method that controls the FWER (i.e.,
Holm--Bonferroni method, \citeproc{ref-holm1979}{Holm, 1979}), and two
cluster-based permutation methods (permutation with a single
cluster-forming threshold and threshold-free cluster enhancement,
\texttt{TFCE}, \citeproc{ref-smith2009}{S. Smith \& Nichols, 2009}). The
\texttt{BH95}, \texttt{BY01}, and \texttt{Holm} corrections were applied
to the p-values using the \texttt{p.adjust()} function in \texttt{R}.
The cluster-based inference was implemented using a cluster-sum
statistic of squared \(t\)-values, as implemented in \texttt{MNE-Python}
(\citeproc{ref-gramfort2013}{Gramfort, 2013}), called via the \texttt{R}
package \texttt{reticulate} v 1.42.0 (\citeproc{ref-reticulate}{Ushey et
al., 2024}). We also compared these estimates to the onset and offset as
estimated using the binary segmentation algorithm, as implemented in the
\texttt{R} package \texttt{changepoint} v 2.3
(\citeproc{ref-changepoint}{Killick et al., 2022}), and applied directly
to the squared \(t\)-values (as in
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025}).\footnote{As in
  Rousselet (\citeproc{ref-rousselet_using_2025}{2025}), we fixed the
  number of expected change points to two in the binary segmentation
  algorithm, thus producing always one cluster.}
Figure~\ref{fig-corrections} illustrates the onsets and offsets
estimated by each method on a single simulated dataset and shows that
all methods systematically overestimate the true onset and underestimate
the true offset. In addition, the \texttt{Raw\ p-value},
\texttt{FDR\ BH95}, and \texttt{FDR\ BY01} methods identify clusters
well before the true onset and after the true offset.

@@

\caption{\label{fig-corrections}Exemplary timecourse of squared t-values
with true onset and offset (vertical black dashed lines) and
onsets/offsets identified using the raw p-values, the corrected p-values
(BH95, BY01, Holm), the cluster-based methods (Cluster mass, TFCE), or
using the binary segmentation method (Change point).}

@@

To assess the accuracy of group-level onset and offset estimation, all
methods were compared by computing the bias (defined as the mean
difference between the estimated and true value of the onset/offset),
mean absolute error (MAE), root mean square error (RMSE), and variance
of onset/offset estimates from 10,000 simulated datasets. Following
Rousselet (\citeproc{ref-rousselet_using_2025}{2025}), each participant
was assigned a random onset between 150 and 170ms. Whereas the present
article focuses on one-dimensional signals (e.g., one M/EEG channel), we
provide a 2D application in \hyperref[apx-2D]{Appendix~A}.

@@

To complement the simulation study, we evaluated the performance of all
methods on actual MEG data (\citeproc{ref-nalborczyk:inprep}{Nalborczyk
et al., in preparation}). In this study, the authors conducted
time-resolved multivariate pattern analysis (MVPA, also known as
decoding) of MEG data recorded in 32 human participants during a reading
task. As a result, the authors obtained a timecourse of decoding accuray
(ROC AUC), bounded between 0 and 1, for each participant. To test
whether the group-level average decoding accuracy was above chance
(i.e., 0.5) at each timestep, we fitted a BGAM as introduced previously
with a basis dimension \(k=50\) and retained all timesteps exceeding a
posterior odds of 20. To better distinguish signal from noise, we
defined a region of practical equivalence (ROPE,
\citeproc{ref-kruschke2017}{Kruschke \& Liddell, 2017}) as the upper
90\% quantile of decoding performance during the baseline period (i.e.,
before stimulus onset). Although we chose a basis dimension of \(k=50\),
which seemed appropriate for the present data, this choice should be
adapted according to the properties of the modelled data (e.g.,
signal-to-noise ratio, prior low-pass filtering, sampling rate) and
should be assessed by the usual model checking tools (e.g., models
comparison, posterior predictive checks, see
\hyperref[apx-basis]{Appendix~B}).

@@

We then assessed the sensitivity of the various methods using a form of
permutation-based sensitivity study, which consisted of the following
steps. First, we created a large number of split halves of the data,
that is, subsets of the dataset containing only 16 out of 32
participants. For each possible pair of subsets, we have 16 possible
levels of overlap/similarity that can be quantified using the Jaccard
index, ranging from 0 (perfectly disjoints subsets) to \(\approx0.88\)
(identical subsets except one participant). For each of these 16 levels
of Jaccard similarity, we created 1,000 pairs of subsets, resulting in
16,000 pairs of subsets in total. For each of these pairs, we estimated
the onset and offset according to each method and computed the absolute
difference in onset/offset estimates. Finally, we estimated the
Spearman's rank correlation coefficient (which quantifies the strength
of a monotonic relation between two variables) between the Jaccard
similarity and the absolute difference in onset/offset estimates. The
rational for this procedure is that sensitive methods should produce
similar onset/offset estimates for similar subsets and dissimilar
onset/offset estimates for dissimilar subsets. The results of this
procedure are summarised in Figure~\ref{fig-splits-similarity}.

@@

Figure~\ref{fig-choose-k} presents the posterior predictions and two
forms of posterior predictive checks (PPCs) for each GAM fit using
different numbers of basis functions (\(k \in \{5, 10, 20, 40, 80\}\)).
With the exception of the \(k=5\) model, all other fits yield
satisfactory PPCs, indicating that the predicted data closely resemble
the empirical observations. However, model comparison using the
leave-one-out information crtierio (LOOIC), as summarised in
Table~\ref{tbl-simulation-results}, identifies the \(k=40\) model as the
best-performing one in terms of LOOIC, closely followed by the \(k=80\)
model. This suggests that the optimal number of basis functions likely
lies between these two values. Future simulation studies could further
investigate how such model selection criteria relate to the precision of
onset and offset estimates.

@@

\CommentTok{\# using the testing\_through\_time() function from the neurogam package}
\CommentTok{\# this may take a few minutes (or hours depending on the machine\textquotesingle{}s}
\CommentTok{\# performance and the size of the dataset)...}
\NormalTok{gam\_onset\_offset }\OtherTok{\textless{}{-}} \FunctionTok{testing\_through\_time}\NormalTok{(}
    \CommentTok{\# dataframe with M/EEG data in long format}
    \AttributeTok{data =}\NormalTok{ raw\_df,}
    \CommentTok{\# the *\_id arguments are used to specify the relevant columns in data }
    \AttributeTok{participant\_id =} \StringTok{"participant"}\NormalTok{, }\AttributeTok{meeg\_id =} \StringTok{"eeg"}\NormalTok{,}
    \AttributeTok{time\_id =} \StringTok{"time"}\NormalTok{, }\AttributeTok{predictor\_id =} \StringTok{"condition"}\NormalTok{,}
    \CommentTok{\# posterior odds threshold for defining clusters (20 by default)}
    \AttributeTok{threshold =} \DecValTok{20}\NormalTok{,}
    \CommentTok{\# number of warmup MCMC iterations}
    \AttributeTok{warmup =} \DecValTok{1000}\NormalTok{,}
    \CommentTok{\# total number of MCMC iterations}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \CommentTok{\# number of MCMCs}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \CommentTok{\# number of parallel cores to use for running the MCMCs}
    \AttributeTok{cores =} \DecValTok{4}
\NormalTok{    )}

@@

\CommentTok{\# activating automatic pandas{-}R conversion}
\NormalTok{pandas2ri.activate()}

>>>>

\abstract{Time-resolved electrophysiological measurements such as those
obtained through magneto- and electroencephalography (M/EEG) offer a
unique window onto the neural activity underlying cognitive processes.
Researchers are often interested in determining whether and when these
signals differ across experimental conditions or participant groups. The
conventional approach involves mass univariate statistical testing
across time and space followed by corrections for multiple comparisons
or some form of cluster-based inference. While effective for controlling
error rates at the cluster-level, cluster-based inference comes with a
significant limitation: by shifting the focus of inference from
individual time points to clusters, it prevents drawing conclusions
about the precise onset or offset of observed effects. Here, we present
a \emph{model-based} alternative for analysing M/EEG timeseries, such as
event-related potentials or time-resolved decoding accuracy. Our
approach leverages Bayesian generalised additive multilevel models,
providing posterior odds that an effect exceeds zero (or chance) at each
time point, while naturally accounting for temporal dependencies and
between-subject variability. Using both simulated and empirical M/EEG
datasets, we show that this approach substantially outperforms
conventional methods in estimating the onset and offset of neural
effects, yielding more precise and reliable estimates. We provide an
open-source R package implementing the method and describe how it can be
integrated into M/EEG analysis pipelines using MNE-Python. }

@@

Understanding the temporal dynamics of cognitive processes requires
methods that can capture fast-changing neural activity with high
temporal resolution. Magnetoencephalography and electroencephalography
(M/EEG) are two such methods, widely used in cognitive neuroscience for
their ability to track brain activity at the millisecond scale. These
techniques provide rich timeseries data that reflect how neural
responses unfold in response to stimuli or tasks. A central goal in many
M/EEG studies is to determine whether, when, and where neural responses
differ across experimental conditions or groups.

The conventional approach involves mass univariate statistical testing
through time and/or space followed by some form of correction for
multiple comparisons with the goal of maintaining the family-wise error
rate (FWER) or the false discovery rate (FDR) at the nominal level
(e.g., 5\%). Cluster-based inference is the most common way of achieving
this sort of error control in the M/EEG literature, being the
recommended approach in several software programs (e.g.,
\texttt{EEGlab}, \citeproc{ref-delorme2004}{Delorme \& Makeig, 2004};
\texttt{MNE-Python}, \citeproc{ref-gramfort2013}{Gramfort, 2013}). While
effective for controlling error rates, cluster-based inference comes
with a significant limitation: by shifting the focus of inference from
individual datapoints (e.g., timesteps, sensors, voxels) to clusters, it
prevents the ability to draw precise conclusions about the
spatiotemporal localisation of such effects
(\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}). As
pointed out by Maris \& Oostenveld (\citeproc{ref-maris2007}{2007}):
``there is a conflict between this interest in localized effects and our
choice for a global null hypothesis: by controlling the FA {[}false
alarm{]} rate under this global null hypothesis, one cannot quantify the
uncertainty in the spatiotemporal localization of the effect''. Even
worse, Rosenblatt et al. (\citeproc{ref-rosenblatt2018}{2018}) note that
cluster-based inference suffers from low spatial resolution: ``Since
discovering a cluster means that `there exists at least one voxel with
an evoked response in the cluster', and not that `all the voxels in the
cluster have an evoked response', it follows that the larger the
detected cluster, the less information we have on the location of the
activation.'' As a consequence, cluster-based inference is expected to
perform poorly for identifying the onset of M/EEG effects; a property
that was later demonstrated in simulation studies (e.g.,
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}).

@@

Scalp-recorded M/EEG signals capture neural activity originating from
various brain regions and are often contaminated by artefacts unrelated
to the cognitive processes under investigation. Consequently, analysing
M/EEG data necessitates methods that can disentangle task-relevant
neural signals from extraneous ``noise.'' A widely adopted technique for
this purpose is the estimation of event-related potentials (ERPs), which
are stereotyped electrophysiological responses time-locked to specific
sensory, cognitive, or motor events. Typically, ERPs are derived by
averaging EEG or MEG epochs across multiple trials aligned to the event
of interest (e.g., stimulus onset), thereby enhancing the
signal-to-noise ratio by attenuating non-time-locked activity. However,
this averaging approach has notable limitations: it assumes consistent
latency and amplitude across trials and is primarily suited for simple
categorical designs. Such assumptions may not hold in more complex
experimental paradigms, potentially leading to suboptimal ERP
estimations (e.g., \citeproc{ref-smith2014a}{N. J. Smith \& Kutas,
2014a}).

To overcome the limitations of simple averaging, several model-based
approaches for estimating ERPs have been proposed. These methods are
motivated by the observation that traditional ERP averaging is
mathematically equivalent to fitting an intercept-only linear regression
model in a simple categorical design without overlapping events
(\citeproc{ref-smith2014a}{N. J. Smith \& Kutas, 2014a}). In contrast to
simple averaging, regression-based approaches to ERP estimation offer
substantially greater flexibility. Notably, they allow for the modelling
of both linear and nonlinear effects of continuous predictors, such as
word frequency or age (e.g., \citeproc{ref-smith2014a}{N. J. Smith \&
Kutas, 2014a}, \citeproc{ref-smith2014b}{2014b};
\citeproc{ref-tremblay2014}{Tremblay \& Newman, 2014}), and enable the
disentangling of overlapping cognitive processes (e.g.,
\citeproc{ref-ehinger_unfold_2019}{Ehinger \& Dimigen, 2019};
\citeproc{ref-skukies_brain_2024}{Skukies et al., 2024};
\citeproc{ref-skukies_modelling_2021}{Skukies \& Ehinger, 2021}). One
widely used implementation of this approach is provided by the LIMO EEG
toolbox (\citeproc{ref-pernet2011}{Pernet et al., 2011}), which follows
a multi-stage analysis pipeline. First, a separate regression model is
fit for each datapoint (e.g., each time point and electrode) at the
individual level to estimate ERP responses. This is followed by
group-level statistical analyses of the resulting regression
coefficients, often accompanied by corrections for multiple comparisons
or cluster-based inference (for recent applied examples, see
\citeproc{ref-dunagan2025}{Dunagan et al., 2025};
\citeproc{ref-wuxfcllhorst2025}{Wüllhorst et al., 2025}).

@@

Beyond modelling nonlinear effects of continuous predictors on ERP
amplitudes, GAMs have been employed to capture the temporal dynamics of
ERPs themselves, effectively modelling the shape of the waveform over
time (\citeproc{ref-abugaber2023}{Abugaber et al., 2023};
\citeproc{ref-baayen2018}{Baayen et al., 2018};
\citeproc{ref-meulman2015}{Meulman et al., 2015},
\citeproc{ref-meulman2023}{2023}). This approach allows for the
estimation of smooth, data-driven functions that characterise how neural
responses evolve over time, offering a flexible alternative to
traditional linear models. In the following section, we provide a brief
introduction to GAMs, highlighting their applicability to M/EEG
timeseries analysis and the advantages they offer over conventional
methods.

@@

A detailed treatment of the technical underpinnings of GAMs is beyond
the scope of this article (see reference books such as
\citeproc{ref-hastie2017}{Hastie \& Tibshirani, 2017};
\citeproc{ref-wood2017}{Wood, 2017a}). However, it is worth emphasising
that GAMs have been successfully applied to a wide range of timeseries
data across the cognitive sciences, including pupillometry (e.g.,
\citeproc{ref-vanrij2019}{Rij et al., 2019}), articulography (e.g.,
\citeproc{ref-wieling2018}{Wieling, 2018}), speech formant dynamics
(e.g., \citeproc{ref-suxf3skuthy2021}{Sóskuthy, 2021}), neuroimaging
data (e.g., \citeproc{ref-dinga2021}{Dinga et al., 2021}), and
event-related potentials (e.g., \citeproc{ref-abugaber2023}{Abugaber et
al., 2023}; \citeproc{ref-baayen2018}{Baayen et al., 2018};
\citeproc{ref-meulman2015}{Meulman et al., 2015},
\citeproc{ref-meulman2023}{2023}). Their appeal for modelling M/EEG data
lies in their ability to flexibly capture the complex shape of ERP
waveforms without overfitting, through the use of smooth functions
constrained by penalisation. Recent extensions, such as distributional
GAMs (\citeproc{ref-rigby2005}{Rigby \& Stasinopoulos, 2005};
\citeproc{ref-umlauf2018}{Umlauf et al., 2018}), allow researchers to
model not only the mean structure but also the variance (or scale) and
other distributional properties as functions of predictors, a feature
that has proven useful in modelling neuroimaging data (e.g.,
\citeproc{ref-dinga2021}{Dinga et al., 2021}). Moreover, hierarchical or
multilevel GAMs (\citeproc{ref-pedersen_hierarchical_2019}{E. J.
Pedersen et al., 2019}) provide a principled way to account for the
nested structure of M/EEG data (e.g., trials within participants),
enabling the inclusion of varying intercepts, slopes, and smoothers (as
illustrated in Figure~\ref{fig-intro-gam}\ignorespaces C-D). This
approach mitigates the risk of overfitting and reduces the influence of
outliers on smooth estimates (\citeproc{ref-baayen2020}{Baayen \& Linke,
2020}; \citeproc{ref-meulman2023}{Meulman et al., 2023}).

@@

We then compared the ability of the BGAM to accurately estimate the
onset and offset of the ERP difference to other widely-used methods.
First, we conducted mass univariate t-tests (thus treating each timestep
independently) and identified the onset and offset of the ERP difference
as the first and last values crossing an arbitrary significance
threshold (\(\alpha = 0.05\)). We then followed the same approach but
after applying different forms of multiplicity correction to the
\(p\)-values. We compared two methods that control the FDR (i.e.,
\texttt{BH95}, \citeproc{ref-benjamini1995}{Benjamini \& Hochberg,
1995}; and \texttt{BY01}, \citeproc{ref-benjamini2001}{Benjamini \&
Yekutieli, 2001}), one method that controls the FWER (i.e.,
Holm--Bonferroni method, \citeproc{ref-holm1979}{Holm, 1979}), and two
cluster-based permutation methods (permutation with a single
cluster-forming threshold and threshold-free cluster enhancement,
\texttt{TFCE}, \citeproc{ref-smith2009}{S. Smith \& Nichols, 2009}). The
\texttt{BH95}, \texttt{BY01}, and \texttt{Holm} corrections were applied
to the \(p\)-values using the \texttt{p.adjust()} function in
\texttt{R}. The cluster-based inference was implemented using a
cluster-sum statistic of squared \(t\)-values, as implemented in
\texttt{MNE-Python} (\citeproc{ref-gramfort2013}{Gramfort, 2013}),
called via the \texttt{R} package \texttt{reticulate} v 1.42.0
(\citeproc{ref-reticulate}{Ushey et al., 2024}). We also compared these
estimates to the onset and offset estimated using the binary
segmentation algorithm, as implemented in the \texttt{R} package
\texttt{changepoint} v 2.3 (\citeproc{ref-changepoint}{Killick et al.,
2022}), and applied directly to the squared \(t\)-values (as in
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025}).\footnote{As in
  Rousselet (\citeproc{ref-rousselet_using_2025}{2025}), we fixed the
  number of expected change points to two in the binary segmentation
  algorithm, thus producing always one cluster.}
Figure~\ref{fig-corrections} illustrates the onsets and offsets
estimated by each method on a single simulated dataset and shows that
all methods systematically overestimate the true onset and underestimate
the true offset. In addition, the \texttt{Raw\ p-value},
\texttt{FDR\ BH95}, and \texttt{FDR\ BY01} methods identify clusters
well before the true onset and after the true offset.

@@

\caption{\label{fig-corrections}Exemplary timecourse of squared t-values
with true onset and offset (vertical black dashed lines) and clusters
identified using the raw p-values, the corrected p-values (BH95, BY01,
Holm), the cluster-based methods (Cluster mass, TFCE), or using the
binary segmentation method (Change point).}

@@

To assess the accuracy of group-level onset and offset estimation, all
methods were compared by computing the bias (defined as the mean
difference between the estimated and true value of the onset/offset),
mean absolute error (MAE), root mean square error (RMSE), and variance
of onset/offset estimates from 10,000 simulated datasets. Following
Rousselet (\citeproc{ref-rousselet_using_2025}{2025}), each participant
was assigned a random onset between 150 and 170ms. Whereas the present
article focuses on one-dimensional signals (e.g., one M/EEG channel), we
provide an application to 2D temporal data in
\hyperref[apx-2D]{Appendix~A}.

@@

To complement the simulation study, we evaluated the performance of all
methods on actual MEG data (\citeproc{ref-nalborczyk:inprep}{Nalborczyk
et al., in preparation}). In this study, the authors conducted
time-resolved multivariate pattern analysis (MVPA, also known as
decoding) of MEG data recorded in 32 human participants during a reading
task. As a result, the authors obtained a timecourse of decoding
accuracy (ROC AUC), bounded between 0 and 1, for each participant. To
test whether the group-level average decoding accuracy was above chance
(i.e., 0.5) at each timestep, we fitted a BGAM as introduced previously
with a basis dimension \(k=50\) and retained all timesteps exceeding a
posterior odds of 20. To better distinguish signal from noise, we
defined a region of practical equivalence (ROPE,
\citeproc{ref-kruschke2017}{Kruschke \& Liddell, 2017}) as the upper
90\% quantile of decoding performance during the baseline period (i.e.,
before stimulus onset). Although we chose a basis dimension of \(k=50\),
which seemed appropriate for the present data, this choice should be
adapted according to the properties of the modelled data (e.g.,
signal-to-noise ratio, prior low-pass filtering, sampling rate) and
should be assessed by the usual model checking tools (e.g., models
comparison, posterior predictive checks, see
\hyperref[apx-basis]{Appendix~B}).

@@

We then assessed the sensitivity of all methods using a form of
permutation-based sensitivity study, which consisted of the following
steps. First, we created a large number of split halves of the data,
that is, subsets of the dataset containing only 16 out of 32
participants. For each possible pair of subsets, we have 16 possible
levels of overlap that can be quantified using the Jaccard index,
ranging from 0 (perfectly disjoints subsets) to \(\approx0.88\)
(identical subsets except one participant). For each of these 16 levels
of Jaccard similarity, we created 1,000 pairs of subsets, resulting in
16,000 pairs of subsets in total. For each of these pairs, we estimated
the onset and offset according to each method and computed the absolute
difference in onset/offset estimates. Finally, we estimated the
Spearman's rank correlation coefficient (which quantifies the strength
of a monotonic relation between two variables) between the Jaccard
similarity and the absolute difference in onset/offset estimates. The
rational for this procedure is that sensitive methods should produce
similar onset/offset estimates for similar subsets and dissimilar
onset/offset estimates for dissimilar subsets. The results of this
procedure are summarised in Figure~\ref{fig-splits-similarity}.

@@

Figure~\ref{fig-choose-k} presents the posterior predictions and two
forms of posterior predictive checks (PPCs) for each GAM fit using
different numbers of basis functions (\(k \in \{5, 10, 20, 40, 80\}\)).
With the exception of the \(k=5\) model, all other fits yield
satisfactory PPCs, indicating that the predicted data closely resemble
the empirical observations. However, model comparison using the
leave-one-out information criterion (LOOIC), as summarised in
Table~\ref{tbl-simulation-results}, identifies the \(k=40\) model as the
best-performing one in terms of LOOIC, closely followed by the \(k=80\)
model. This suggests that the optimal number of basis functions likely
lies between these two values. Future simulation studies could further
investigate how such model selection criteria relate to the precision of
onset and offset estimates.

@@

\CommentTok{\# using the testing\_through\_time() function from the neurogam package}
\CommentTok{\# this may take a few minutes (or hours depending on the machine\textquotesingle{}s}
\CommentTok{\# performance and the size of the dataset)...}
\NormalTok{gam\_onset\_offset }\OtherTok{\textless{}{-}} \FunctionTok{testing\_through\_time}\NormalTok{(}
    \CommentTok{\# dataframe with M/EEG data in long format}
    \AttributeTok{data =}\NormalTok{ df,}
    \CommentTok{\# the *\_id arguments are used to specify the relevant columns in data }
    \AttributeTok{participant\_id =} \StringTok{"participant"}\NormalTok{, }\AttributeTok{meeg\_id =} \StringTok{"eeg"}\NormalTok{,}
    \AttributeTok{time\_id =} \StringTok{"time"}\NormalTok{, }\AttributeTok{predictor\_id =} \StringTok{"condition"}\NormalTok{,}
    \CommentTok{\# posterior odds threshold for defining clusters (20 by default)}
    \AttributeTok{threshold =} \DecValTok{20}\NormalTok{,}
    \CommentTok{\# number of warmup MCMC iterations}
    \AttributeTok{warmup =} \DecValTok{1000}\NormalTok{,}
    \CommentTok{\# total number of MCMC iterations}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \CommentTok{\# number of MCMCs}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \CommentTok{\# number of parallel cores to use for running the MCMCs}
    \AttributeTok{cores =} \DecValTok{4}
\NormalTok{    )}

@@

\CommentTok{\# activating automatic pandas{-}R conversion}
\CommentTok{\# pandas2ri.activate()}
